Loyalty's Limits: The Impact of Postmodernism and Political Polarization on Military Professionalism  (Prospectus)

### 2020-10-22 code/prospectusv1.Rmd
commit

  _This dissertation seeks to explore the impact of increasing political polarization on the professional identity of military officers and the military institution. I develop a theory that argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession, and that such a strain is problematic for an institution such as the military that holds the virtue of loyalty in high regard, and by statutes possesses multiple, simultaneous loyalties: to the Constitution of the United States, to the President qua Commander-in-Chief, to Congress, to the American People, to one's fellow service members, and to one's self. Balancing these competing loyalties becomes more difficult as political polarization worsens, I argue, because of the fundamental and core values - not mere policy differences - that are at the root of many contemporary polarized political battles. Employing a mixed-methods approach that incorporates quantitative analysis and several historical case studies, this dissertation argues that the challenges to conceptions of loyalty among members of the military institution brought about by political polarization have the potential to impact normative concepts of military professionalism, which could be problematic not only for the character and identity of the military as an institution, but for American national security at large._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic? ^ [I believe we have already seen such instances of exactly this situation playing out over the past several years as political polarization has worsened.As examples, consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what be believed to be behavior that may have been illegal by the President regarding aid to Ukraine, or the case of Navy Captain Brett Crozier, who leaked a letter to the press about how his pleas for help in battling the coranavirus aboard the ship he commanded, the USS Theodore Roosevelt. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, themselves, and perhaps more.]

@@

My intended topic is different than these other, still very important topics in which disagreements between the executive and the military occur during wartime. It is my assessment that polarization is causing conflict, or certainly has the potential to do so, between the military and civilian leaders, but in most instances, the subject is about topics other than the use of military force, which is somewhat puzzling, given the military's reason d'etre. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, outside the bounds of what constitutes appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] While I believe that such developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and how.

@@

## How Polarization Impacts Loyalty as a Concept

In this section of the prospectus, I will lay out an initial argument, as well as accompanying hypotheses, which I intend to explore and test in my dissertation. Figure 2, below, shows an initial conception of the linkages between several of the variables discussed in the previous section. From left to right, this conceptual model suggests that worsening political polarization is driving changes in both the institutional credibility of the military (how society views the military) and changes in the conceptions of loyalty and identity (within the military). These two independent variables are, I suggest, driving subsequently driving changes in the concept of military professionalism, the main dependent variable of analysis in my dissertation. Following Huntington's model, I also suggest that changes in military professionalism can impact a variety of outcomes, by which I mean the very same outcomes that other scholars do and have already explored, such as effectiveness on the battlefield [@reiter_democracies_2002; @biddle_democracy_2004] and achieving strategic effectivness [@brooks_creating_2007; @brooks_shaping_2008]. While I intend to examine the broad system of variables shown in Figure 2, the focus of my dissertation, as well as the least-explored linkage in this model, is the relationship between political polarization and changes in the concept of loyalty and identity, and how such changes manifest themselves in the variation of military professionalism. For emphasis and clarity, I do not think that the other link - that between political polarization and institutional credibility - is unimportant. Rather, this link seems to be the focus of more-recent scholarship [@robinson_michael_danger_2018].

@@

A primary contention from the system visualized above is that as political polarization in the domestic sphere increases, the military faces increased challenges in how both individual military leaders as well as the institution conceptualizes, adopts, and exercises loyalty, and by extension, professional identity. This expectation stems from several features that scholars of American politics have raised regarding the current politically polarized climate within the United States. First, scholars have noted that political polarization raises the number and scope of issues that are subsumed by partisan debate and disagreement (a phenomenon referred to among scholars of American Politics as "conflict extension" [@layman_party_2002]). A second feature of contemporary political polarization is the prominence of  negative feelings held by one partisan group towards another, a phenomenon referred to as "negative partisanship" [@abramowitz_great_2018]. A third feature is an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. I contend that these features impact individuals serving in the military, including military leaders, as they do members of the domestic public writ large, and that these impacts are problematic for members of an institution that normatively exists as an apolitical and non-partisan actor. Table 1, below, lays out how these three prominent features of domestic political polarilzation may impact the non-partisan and apolitical nature of the normative military professional and institution.

@@

The first way in which the apolitical and non-partisan ethos of the military is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This happens because as "conflict extension" [@layman_party_2002] occurs, an increasing number of issues are brought into partisan battlegrounds, and thus, there is simply less politically-neutral territory available. Thus, an actor who wishes to remain non-partisan, such as the military, may be increasingly less able to do so, as a result of this feature.

A second challenge military leaders and the institution faces in a polarized climate is from the "negative partisanship"[@abramowitz_great_2018] that dominates the polarized American political landscape. As citizens hold increasingly strong and negative feelings towards those who side differently on a host of issues, there may be a tendency for members of the military institution to harbor similar feelings. Even if impact of this feature on the military and its leaders, due to training and professional development that emphasizes the importance of upholding apolitical and non-partisan behavioral norms, is somewhat muted, the point is that members of the military institution are likely experiencing the same phenomenon - even if it is to a lesser degree - as the country's citizens.

A third, and I contend the most impactful, challenge that normative military professionalism faces stems from the belief by individuals that "core" issues are at stake in the current polarized political climate. "Core" issues are difficult to define precisely [need cite], but the main point is that these issues often involve perceptions of moral right and wrong, and not just simple policy preferences. Moreover, some of the most polarizing issues are often associated with issues related to an individual's identity, such as race, gender, immigration [need cite]. The rise of "identity politics" in America as political polarization has worsened illustrates this point [need cite].

I argue that this third challenge is especially relevant for the concept of loyalty. As an increasing number of what individuals believe to be "core" issues are raised and debated in the partisan arena, individuals face decisions of how to prioritize the persons or objects of their loyalty and further, what actions to take on the basis of loyalty to these persons or objects. I further theorize (though this has to be tested, of course) that while all individuals will likely face some version of this challenge of loyalty in a polarizing age, the challenges will be more acute for military leaders. Not only is the de-facto setting in which military leaders operate marked by multiple and often conflicting loyalties due to the structure of and statutes set forth in the US Constitution, but also, the potential consequences are higher for military leaders than average citizens should they fail to discharge the duties that correspond to their roles as professional officers. These consequences include failure to achieve tangible national security goals, such as loss on the battlefield, or less tangible but still important outcomes such as maintaining trust with the American public.

Moreover, as important as these consequences are, they do not speak to the potentially high personal and individualistic consequences military leaders may face if confronted with a situation in which they have to execute orders which are technically legal, but morally borderline at best. Military leaders are still humans who have to live with themselves at the end of the day. Therefore, I argue that if polarization continues to erode boundaries of what is considered moral by individuals to begin with, there is an increased likelihood of military leaders having to choose between conflicting loyalties. What happens if and when the objects of a military leader's multiple loyalties not merely disagree, but become stubbornly divided ideologically? Can such an outcome exist, and if it can, is there empirical proof that such a trend is occurring? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

@@

The previous section described how increasing political polarization presents significant challenges to aspects of the normative concept of military professionalism. But whether these challenges enter the public domain, or remain private challenges that specific individual actors contend with on a personal level only, matters a great deal because it "changes the character...from a matter of principle to a political matter" [@dubik_principled_2014] In fact, in much of the field of civil-military relations, this dynamic of private disagreement or discontent potentially entering the public realm is seen in the scholarly literature on senior officer resignation. This debate examines various aspects of resignation, including whether there can ever be conditions under which resignation is justified or appropriate, with scholars taking various sides on the issue, both for [@dubik_principled_2014; @snider_strategic_2014] and against [@feaver_should_2014; @golby_beyond_2015]. I contend that polarization and credibility of the military interact, however, in at least two important ways. First, as loyalties are challenged as a result of increasing political polarization, individual military leaders may not necessarily remain quiet despite the existence of a strong professional norm of doing so because the objects of loyalty may demand that their adherents publicly speak out. This of course, depends specifically on what the object of loyalty is and what it demands from those who hold closely to it. The second way in which polarization and credibility may interact arises more directly from what can be termed a credibility effect, and has more to do with the place of prominence (or lack thereof) that the military or individual leaders hold in the perception of American society. [need to blend this in with recent scholarship and where gaps are - bring in Urben, Golby, Dropp and Feaver, etc., and Robinson.  Also need to figure out where to put my hypotheses]

@@

The final set of conditions exists when political polarization and institutional credibility are high. It is in this world that I contend the American military finds itself in now in 2020. Under these conditions, the military faces dual dangers. High institutional credibility risks the military institution unilaterally acting or another body co-opting the military institution to speak out an a number of issue areas, and simultaneously, the highly polarized environment risks the military advocating for or denouncing partisan sentiments in the name of upholding the very traits it is most adhered to - loyalty and duty. Indeed, this set of conditions comprise a very dangerous world for the military institution and the military professional.

@@

@@

_H4: Following from the above hypothesis, polarized periods will feature events in which governmental actors appeal to the military's sense of loyalty in order to persuade them to adopt a stance on a particular issue area._

_H5: In highly polarized time periods, it is the senior military officers as the "stewards of the profession" [@us_department_of_the_army_adp_2019, p.1-9] whose actions will most influence to what degree and how the military institution changes. If changes occur to the military institution, they are bound to occur within two primary spheres: the military's professional identity, and the military's professional coherence._

# Plan of the Dissertation and the Data Generating Process

To complete my dissertation, I plan on incorporating a mixed-methods approach. In broad conceptual terms, I plan to propose a theory and then test the theory using a combination of quantitative analyses and qualitative case studies. The one sentence thesis of my dissertation is that political polarization significantly impacts how the military thinks about and exercises loyalty, a chief military trait, and that these impacts can and often do result in changes to the military's professional identity and character. I plan to execute a book style project comprised of at least seven chapters.

In the first chapter, I will frame my research question, lay out my research agenda, and review the relevant literature for my project. This will involve synthesizing several of the key aspects and debates within the field of civil-military relations, much in the same way I have done in the first part of this dissertation prospectus. I will focus my attention on introducing and summarizing the concepts of military professionalism, organizational loyalty, and political polarization, and spend time orienting the reader to the debate within scholarly literature regarding a Huntingtonian conception of military professionalism. I will emphasize the long-running nature of this debate going all the way back to when Huntington first published his book, but point to some of the prominent contemporary critiques regarding his theory [@feaver_civil-military_1996-1; @rapp_civil-military_2015; @golby_why_2017; @karlin_case_2020; @brooks_paradoxes_2020]. The main point of chapter 1 is to set the stage for my dissertation and signal to the reader that modern critiques of Huntington may be overlooking a key part of his concept of military professionalism: it may insulate the military from the most pernicious effects of polarizing politics.

In the second chapter, I will introduce, develop, and explain my theory that links increasing political polarization to conceptions of loyalty and military professionalism (see Figure 2). I will argue that specific features of the polarized environment strain and challenge both individual and institutional conceptions of loyalty, and by extension, normative aspects of military professionalism (Table 1). I will further argue that rising political polarization can interact with changes in the military's institutional credibility to increase the likelihood that the public becomes aware of changes in the military's professional conduct. Using Table 2 as a framework, I will place epochs of historical American civil-military relations into one of four distinct quadrants, and signal that these quadrants will comprise the qualitative case studies that I later examine in my dissertation.

In the third chapter, I will argue that today, in 2020, the American public is witnessing a decline in professional military conduct among retired military officers. I plan to demonstrate this empirically by drawing upon an original dataset consisting of more than 340 op-ed pieces authored by retired military officers since the late 1970s and published in the top five major print newspapers in the United States, including the Wall Street Journal, the New York Times, the Washington Post, the USA Today, and the Los Angeles Times. I believe that I can show several things from this analysis, including that there has been a sharp uptick in the number of op-eds published by retired military officers during this time period, and more importantly, that there has been a decline in the normative professional ideals espoused and argued for in these op-eds. Critical to this chapter is acknowledgement that coding for what is professional versus what is not is subjective to a large extent, and thus, the need for me to truly develop clear working definitions of the concept of military professionalism in Chapter 1. However, I will show that whether one adopts a strict Huntingtonian version of military professionalism, or a less strict version that allows for more encroachment by the military institution into the political sphere, by both standards the retired military officer corps is not living up to the ideals of normative military professionalism. The question, then, becomes why aren't they? I will argue that while there are certainly other confounding variables, to include changes in the conservative-liberal tilt of newspapers over time, the chief explanation for the drop in professionalism stems from the combined effects of increasing political polarization and rising institutional credibility of the military since the late 1970s, and the challenges that these changes introduce to a military leader's conception of loyalty. Moreover, my ultimate point is to point out a blatant inconsistency in the way that the military currently thinks about the political behavior by those within the ranks in that the institution has rules that governs the conduct of those on active duty [@department_of_defense_political_2008], but those same rules, and the spirit behind those rules, are disregarded by the institution's vangaurds, its retired multistar officers and "stewards of the profession" [@us_department_of_the_army_adp_2019, p. 1-9] of arms.

In the fourth chapter, I want to show that the institutional challenges to loyalty have been present as political polarization has worsened. To do this, I plan to examine, via textual analysis, the records of Congressional Committee hearings in which military officers, active and retired, have played a part, and detect if and how patterns of questioning and providing answers have changed over time. I am especially interested in the way that individuals have invoked language associated with the concept of loyalty in asking and answering questions. An example of this might be analyzing the ways in which terms such as "moral obligation" or similar language is used in these hearings. I hope to detect patterns that will help support my overall argument. I have yet to flesh this particular chapter out some more.

In Chapter 5, I plan to analyze the results of an original survey experiment in which I survey a nationally representative sample of civilians and compare the results to a sample of either a) veterans, or b) active duty military officers. In this survey experiment, I plan to test whether members of the military (which I proxy by veterans) genuinely are more sensitive to appeals to loyalty, as I have hypothesized in _H3._ The survey, after controlling for gender, partisan identification, and age, will present the respondent with 6-8 hypothetical situations and a corresponding argument made by a military elite. Members of the treatment group will receive an appeal to loyalty as part of the elite's argument as to why a certain course of action should be followed regarding the hypothetical situation presented.

@@

\doublespace
Such a survey will enable me to measure two things. First, it will allow me to measure whether appeals to loyalty are truly more effective than standard appeals to action. This result can be measured by comparing the answers between the treatment and control groups. Second, it will allow me to measure whether members of the military (or veterans) react differently to these appeals to loyalty than regular civilians. As I have hypothesized in _H3_, I suspect that they do. Of note, I am planning to also pursue a survey among actively-serving military officers, but the lead time for obtaining approval to do this can be very long, perhaps 9-12 months or more. If I am able to obtain approval for this, I will then have three different populations to measure: civilians, veterans, and actively-serving military officers. Of note, if I discover throughout the course of my survey that loyalty does not in fact have the strong explanatory power that I think it does, or if veterans and/or active military officers do not respond to appeals to loyalty any differently than civilians, I think such a finding is also interesting.

In the sixth chapter, I will turn to a series of qualitative case studies to explore and test my theory. Key in this chapter is finding and examining historical periods in American civil-military relations in which there existed varying levels of political polarization and institutional credibility, and then, to walk the reader through how challenges of loyalty during these periods manifested themselves in different ways among specific members of the military and the institional military at large, in line with the theory that I propose in Chapter 2. My initial projection of what these case studies will be and how they will align with varying levels of political polarization and instititional credibility are as follows:

@@

* High Polarization, Low institutional credibility: civil-military relations in the post-Vietnam period, in the late 1970s.

@@

* Low Polarization, Low Institutional Credibility: TBD.

>>>>

  _This dissertation seeks to explore the impact of increasing political polarization on the professional identity of military officers and the military institution. I develop a theory that argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession, and that such a strain is problematic for an institution such as the military that holds the virtue of loyalty in high regard, and by statute possesses multiple, simultaneous loyalties: to the Constitution of the United States, to the President qua Commander-in-Chief, to Congress, to the American People, to one's fellow service members, and to one's self. Balancing these competing loyalties becomes more difficult as political polarization worsens, I argue, because of the fundamental and core values - not mere policy differences - that are at the root of many contemporary polarized political battles. Employing a mixed-methods approach that incorporates quantitative analysis and several historical case studies, this dissertation argues that the challenges to conceptions of loyalty among members of the military institution brought about by political polarization have the potential to impact normative concepts of military professionalism, which could be problematic not only for the character and identity of the military as an institution, but for American national security at large._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic?^[Loyalty had to be a prime consideration in several prominent, recent cases involving the military and the government. Consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what Vindman believed to be behavior by the President that may have been illegal, or the case of Navy Captain Brett Crozier, who leaked a letter to the press after his pleas for help in battling the coranavirus aboard the ship he commanded, the USS Theodore Roosevelt, went unheeded for what he felt was too long. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, the operational mission, themselves as individual moral agents, and perhaps even more objects of loyalty.]

@@

My intended research topic is different than these other, still very important topics in which disagreements between the executive and the military occur during wartime. It is my assessment that polarization is causing conflict, or certainly has the potential to do so, between the military and civilian leaders, but in most instances, the subject is about topics other than the use of military force, which is somewhat puzzling, given the military's reason d'etre. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, outside the bounds of what constitutes appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation]

This raises, then, an important question: does the military's essential purpose risk evolving or changing significantly during polarized times? While I believe that these potential developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and in particular, the role of the military institution itself in facilitating these developments. Most scholars acknowledge the potentially damaging impacts of the politicization of the military [@golby_listening_2013, @brooks_paradoxes_2020] at the hands of civilian leaders, and I share these concerns. Fewer scholars, if any, are exploring the potential role of the military in politicizing itself, and the mechanisms through which these developments occur.

@@

## Political Polarization Impacts Conceptions of Loyalty

In this section of the prospectus, I will lay out an initial argument, as well as accompanying hypotheses, which I intend to explore and test in my dissertation. Figure 2, below, shows an initial conception of the linkages between several of the variables discussed in the previous section of the prospectus. From left to right, this conceptual model suggests that worsening political polarization is driving changes in both the institutional credibility of the military (how society views the military) and changes in the way members of the military institution think about and exercise loyalty, and by extension, how they think of their professional identity. These two independent variables are, I suggest, driving subsequent changes in the concept of military professionalism, the main dependent variable in my analysis.

In line with Huntington's model as well as a host of other scholarship, I suggest that changes in military professionalism can impact a variety of military outcomes, varying from effectiveness on the battlefield [@reiter_democracies_2002; @biddle_democracy_2004] to achieving strategic effectivness [@brooks_creating_2007; @brooks_shaping_2008]. While I intend to examine the broad system of variables shown in Figure 2, the focus of my dissertation is the relationship between political polarization and changes in the concept of loyalty and identity, and how such changes manifest themselves in the variation of military professionalism. For emphasis and clarity, I do not think that the other link - that between political polarization and institutional credibility - is unimportant. Rather, this link seems to have been the focus of several recent scholarship efforts [@robinson_michael_danger_2018; ].

@@

A primary contention from the system visualized above is that as political polarization in the domestic sphere increases, the military faces increased challenges in how both individual military leaders as well as the institution conceptualizes, adopts, and exercises loyalty, and by extension, how it sees its own professional identity. This expectation stems from several features that scholars of American politics have raised regarding the current politically polarized climate within the United States. First, scholars have noted that political polarization raises the number and scope of issues that are subsumed by partisan debate and disagreement (a phenomenon referred to among scholars of American Politics as "conflict extension" [@layman_party_2002]). A second feature of contemporary political polarization is the prominence of  negative feelings held by one partisan group towards another, a phenomenon referred to as "negative partisanship" [@abramowitz_great_2018]. A third feature is an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. I contend that these features impact individuals serving in the military, including military leaders, as they do members of the domestic public writ large, and that these impacts are problematic for members of an institution that normatively exists as an apolitical and non-partisan actor. Table 1, below, lays out how these three prominent features of domestic political polarilzation may impact the non-partisan and apolitical nature of the normative military professional and institution.

@@

The first way in which the apolitical and non-partisan ethos of the military is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This happens because as "conflict extension" [@layman_party_2002] occurs, an increasing number of issues are brought into partisan battlegrounds, and thus, there is simply less politically-neutral territory available. Thus, an actor such as the military, who may wish to remain non-partisan and out of politics to the extent that it can, may simply be less able to do so.

A second challenge military leaders and the institution faces in a polarized climate is from the rise of "negative partisanship"[@abramowitz_great_2018] that dominates the polarized American political landscape. As citizens hold increasingly strong and negative feelings towards those who side differently on a host of issues, there may be a tendency for members of the military institution to harbor similar feelings and attitudes. Even if the impact of this feature on the military and its leaders, due to training and professional development that emphasizes the importance of upholding apolitical and non-partisan behavioral norms is somewhat muted, the point is that members of the military institution are likely experiencing the same phenomenon - even if it is to a lesser degree - as the country's citizens. One important question about "negative partisanship" in a military setting, though, is whether service members hold strong negative feelings toward only civilians who are on the opposite side of the ideological spectrum, or whether such feelings also permeate within the ranks.

A third, and I contend the most impactful, challenge that normative military professionalism faces stems from the belief by individuals that core issues - such as those that involve attitudes about race, religion, and multiple layers of identity - are at stake in the current polarized political climate. "Core" issues are difficult to define precisely, but the main point is that these issues often involve perceptions of identity [] and moral right and wrong, and not just simple policy preferences. Moreover, some of the most polarizing issues are often associated with issues related to an individual's identity, such as race, gender, immigration, as the the rise of "identity politics" in America illustrates this point [@sides_identity_2018].

I argue that this third challenge is especially relevant for the contemporary military, and specifically, with respect to how members of the military exercise the concept of loyalty. As an increasing number of what individuals believe to be core issues are raised and debated in the partisan arena, individuals face decisions of how to prioritize the persons or objects of their loyalty and further, what actions to take on the basis of loyalty to these persons or objects. I further theorize (though this has to be tested, of course) that while all individuals will likely face some version of this challenge of loyalty in a polarizing age, the challenges will be more acute for military leaders. Not only is the de-facto setting in which military leaders operate marked by multiple and often conflicting loyalties due to the structure of and statutes set forth in the US Constitution, but also, the potential consequences are higher for military leaders than average citizens should they fail to discharge the duties that correspond to their roles as professional officers. These consequences include failure to achieve tangible national security goals, such as loss on the battlefield, or less tangible but still important outcomes such as maintaining trust with the American public.

As important as these consequences are, they do not speak to another set of consequences individual military leaders also may face in a polarizing age. Indeed, in addition to consequences to the country that may occur should the military fail to discharge its duties to safeguard America, military leaders also potential consequences if confronted with a situation in which they have to execute orders which are technically legal, but morally borderline at best. Military leaders are still "moral agents" who "remain responsible for the decisions and actions they take" [@dubik_just_2016, 160]. But the considerations of what constitutes moral right and wrong to begin with may be changing as polarization worsens [@graham_liberals_2009].

Therefore, I argue that if polarization continues to heighten the identities of groups that people associate with, as well as come closer to the boundaries of how individuals think about moral right and wrong, there is an increased likelihood of military leaders having to choose between conflicting loyalties. What happens if and when the objects of a military leader's multiple loyalties not merely disagree, but become stubbornly divided ideologically? Can such an outcome exist? Is there empirical proof that such a trend is happening? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

@@

The previous section described how increasing political polarization presents significant challenges to aspects of the normative concept of military professionalism. But whether these challenges enter the public domain, or remain private challenges that specific individual actors contend with on a personal level only, matters a great deal because it "changes the character...from a matter of principle to a political matter" [@dubik_principled_2014] In fact, in much of the field of civil-military relations, this dynamic of private disagreement or discontent potentially entering the public realm is seen in the scholarly literature on senior officer resignation. This debate examines various aspects of resignation, including whether there can ever be conditions under which resignation is justified or appropriate, with scholars taking various sides on the issue, both for [@dubik_principled_2014; @snider_strategic_2014] and against [@feaver_should_2014; @golby_beyond_2015]. I contend that polarization and credibility of the military interact, however, in at least two important ways. First, as loyalties are challenged as a result of increasing political polarization, individual military leaders may not necessarily remain quiet despite the existence of a strong professional norm of doing so because the objects of loyalty may demand that their adherents publicly speak out. This of course, depends specifically on what the object of loyalty is and what it demands from those who hold closely to it. The second way in which polarization and credibility may interact arises more directly from what can be termed a credibility effect, and has more to do with the place of prominence (or lack thereof) that the military or individual leaders hold in the perception of American society. As recent scholarship has shown, the credibility of the military institution in a variety of contexts depends heavily on partisan affiliation [@urben_party_2013; @golby_listening_2013; @golby_jim_elite_2017; @robinson_danger_2018].

@@

The final set of conditions exists when political polarization and institutional credibility are high. It is in this world that I contend the American military finds itself in now in 2020. Under these conditions, the military faces dual dangers. High institutional credibility risks the military institution unilaterally acting or another body co-opting the military institution to speak out an a number of issue areas, and simultaneously, the highly polarized environment risks the military advocating for or denouncing partisan sentiments in the name of upholding the very virtue it is most normatively adhered to - loyalty.

@@

\singlespace

@@

_H4: Following from the above hypothesis, polarized periods will feature events in which governmental actors appeal to the military's sense of loyalty in order to persuade them to adopt a stance on a particular issue area. These can be viewed as "exploits of loyalty."_

_H5: In highly polarized time periods, it is the senior military officers as the "stewards of the profession" [@us_department_of_the_army_adp_2019, p.1-9] whose actions will most influence to what degree and how the military institution reacts to and changes as a result of "exploits of loyalty." If changes occur to the military institution, they are bound to occur to the professional identity of the military, which includes the degree to which it functions as a profession (vice bureaucracy [@snider_once_2012]) and the degree to which it is obedient to its civilian leaders (vice acting as an activist on behalf of a particular cause)._

\doublespace
# Plan of the Dissertation and the Data Generating Process

To complete my dissertation, I plan on incorporating a mixed-methods approach. In broad conceptual terms, I plan to propose a theory and then test the theory using a combination of quantitative analyses and qualitative case studies. The one sentence thesis of my dissertation is that political polarization significantly impacts how the military thinks about and exercises loyalty, a chief military trait, and that these impacts can and often do result in changes to the military's professional identity and character. I plan to execute a book style project comprised of seven chapters.

In the first chapter, I will frame my research question, lay out my research agenda, and review the relevant literature for my project. This will involve synthesizing several of the key aspects and debates within the field of civil-military relations, much in the same way I have done in the first part of this dissertation prospectus. I will focus my attention on introducing and summarizing the concepts of military professionalism, organizational loyalty, and political polarization, and spend time orienting the reader to the debate within scholarly literature regarding a Huntingtonian conception of military professionalism. I will emphasize the long-running nature of this debate going all the way back to when Huntington first published his book, but point to some of the prominent contemporary critiques regarding his theory [@feaver_civil-military_1996-1; @rapp_civil-military_2015; @golby_why_2017; @karlin_case_2020; @brooks_paradoxes_2020]. The main point of chapter 1 is to set the stage for my dissertation and signal to the reader that modern critiques of Huntington may be overlooking a key part of his concept of military professionalism: imperfect as it may be, it may also insulate the military from the most pernicious effects of polarizing politics.

In the second chapter, I will introduce, develop, and explain my theory that links increasing political polarization to conceptions of loyalty and military professionalism (see Figure 2, above). I will argue that specific features of the polarized environment strain and challenge both individual and institutional conceptions of loyalty, and by extension, normative aspects of military professionalism (Table 1). I will further argue that rising political polarization can interact with changes in the military's institutional credibility to increase the likelihood that the public becomes aware of how the military chooses to respond to the dynamics brought on as the military faces challenges to how it conceptualizes and practices loyalty. Using Table 2 as a framework, I will place epochs of historical American civil-military relations into one of four distinct quadrants, and signal that these quadrants will comprise the qualitative case studies that I later examine in my dissertation.

In the third chapter, I will argue that today, in 2020, the American public is witnessing a decline in professional military conduct among retired military officers. I plan to demonstrate this empirically by drawing upon an original dataset consisting of more than 340 op-ed pieces authored by retired military officers since the late 1970s and published in the top five major print newspapers in the United States, including the Wall Street Journal, the New York Times, the Washington Post, the USA Today, and the Los Angeles Times. I believe that I can show several things from this analysis, including that there has been a sharp uptick in the number of op-eds published by retired military officers during this time period, and more relevant to the theory I propose in the dissertation, that these published op-eds reveal a decline in the normative professional ideals of the military institution. Unlike previous recent works which focus on how polarization impacts society's view of the military, this chapter introduces an empirical record of actors within the military institution (retired senior officers) also changing as a result of increasing political polarization.

Critical to this chapter is acknowledgement that coding for what is professional versus what is not is a somewhat subjective process. Thus, I will need to develop and implement a clear working definition of the concept of military professionalism in Chapter 1. However, I believe that I can show in Chapter 3 that whether one adopts a strictly Huntingtonian version of military professionalism (strict separation of political and military spheres) or a less strict version that allows for an increasing shared space, the retired military officer corps is not living up to the non-partisan ethic that the military espouses.

I will argue that while there are certainly other confounding variables, to include changes in the conservative-liberal tilt of newspapers over time, the chief explanation for the variation (decline) in professionalism stems from the combined effects of increasing political polarization and rising institutional credibility of the military since the late 1970s, and the challenges that these changes introduce to a military leader's conception of loyalty. Moreover, my ultimate point is to point out a blatant inconsistency in the way that the military currently thinks about the political behavior by those within the ranks in that the institution has rules that governs the conduct of those on active duty [@department_of_defense_political_2008], but those same rules, and the spirit behind those rules, are disregarded by the institution's vangaurds, its retired multistar officers and "stewards of the profession" [@us_department_of_the_army_adp_2019, p. 1-9] of arms. I believe that this chapter will result in finding support for _H1._

In Chapter four, I plan to analyze the results of an original survey experiment in which I survey a nationally representative sample of civilians and compare the results to a sample of either a) veterans, or b) active duty military officers. In this survey experiment, I plan to test whether members of the military (which I proxy by veterans) genuinely are more sensitive to appeals to loyalty, as I have hypothesized in _H3._ The survey, after controlling for gender, partisan identification, and age, will present the respondent with 6-8 hypothetical situations and a corresponding argument made by a military elite. Members of the treatment group will receive an appeal to loyalty as part of the elite's argument as to why a certain course of action should be followed regarding the hypothetical situation presented.

@@

\doublespace

Such a survey will enable me to measure two things. First, it will allow me to measure whether appeals to loyalty are truly more effective than standard appeals to action. This result can be measured by comparing the answers between the treatment and control groups. Second, it will allow me to measure whether members of the military (or veterans) react differently to these appeals to loyalty than regular civilians. As I have hypothesized in _H3_, I suspect that they do. Of note, I am planning to also pursue a survey among actively-serving military officers, but the lead time for obtaining approval to do this can be very long, perhaps 9-12 months or more. If I am able to obtain approval for this, I will then have three different populations to measure: civilians, veterans, and actively-serving military officers. Of note, if I discover throughout the course of my survey that loyalty does not in fact have the strong explanatory power that I think it does, or if veterans and/or active military officers do not respond to appeals to loyalty any differently than civilians, I think such a finding is also interesting.

In the fifth chapter, I want to show that the military has also faced institutional challenges to the concept of loyalty as political polarization has worsened. To do this, I plan to examine, via textual analysis, the records of Congressional Committee hearings in which military officers, active and retired, have participated. My goal is to detect if and how patterns of questioning and providing answers have changed, particularly in how language related to the concept of loyalty has been used. I still have some work to do in fleshing this out some more, but my initial idea is to examine how the military is both asked and answers questions by members of Congress, and how these questions and answers invoke language that speaks to various loyalty obligations. An example of this might be analyzing the ways in which terms such as "moral obligation" and "sacred duty" have been used over time. I might be able to associate causes or subjects that are associated with this type of language, and correlate these causes and subjects to the level of polarization present in the country, as well as to the partisan identification of the specific member of Congress asking the question. My goal is to detect patterns that will help support my overall argument, and specifically, _H2_.

In the sixth chapter, I will turn to a series of qualitative case studies to explore and test my theory. Key in this chapter is finding and examining historical periods in American civil-military relations in which there existed varying levels of political polarization and institutional credibility, and then, to walk the reader through how challenges of loyalty during these periods manifested themselves in different ways among specific members of the military and the institional military at large, in line with the theory that I propose in Chapter 2 and specifically, _H2_.I then hope to find evidence for how the actions of the military and its leaders during these periods impacted the military's professional identity and character, in line with _H4_ and _H5_.

I have not yet settled on what the specific case studies be, particularly because I am wresting with whether I need to confine my examination to the time period in which the US military has generally been considered a profession (depending on who you ask, this is generally around World War One, or perhaps a little bit earlier, at the tail end of the 19th Century). My initial reaction is that I do not need to confine myself to this period. My initial projection of what these case studies will be and how they will align with varying levels of political polarization and instititional credibility are as follows:

@@

* High Polarization, Low institutional credibility: civil-military relations after World War One, during the 1920s.

@@

* Low Polarization, Low Institutional Credibility: civil-military relations in the post-Vietnam years, roughly 1972-1980.
